

---
title: 'Reproducible Research: Peer Assessment 1'
output:
  html_document:
    keep_md: yes
  pdf_document: default
  
 
---
# Required libraries
This analysis uses the following libraries
```{r Libraries}
#Required libraries
library(plyr)
library(ggplot2)
library("scales")
library("lubridate",quietly = TRUE)
library("stringr")
library("xtable")
```

## Loading and preprocessing the data
First read in the data from the "activity.zip" file in the working directory and convert the data in the "Date"  column into date format
```{r Load Data, echo = TRUE}
setwd("C:/Users/Keith_2/Documents/R/RepData_PeerAssessment1")
alldata<-read.csv(unzip("activity.zip")) 
alldata$date<-as.Date(alldata$date)

options(scipen = 4) #Forces large numbers to be printed in standard form by knitR
```
Here is an example of the data:
```{r example data}
alldata[5000:5007,]
```

Here is a summary of the data
```{r summary of the data, echo = TRUE}

print(summary(alldata))



#calculate percentage of rows that have NA in the steps column
stepna<-round(100* sum(is.na(alldata$steps))/nrow(alldata),1)

#calculate time period over which samples were taken
dur<-max(alldata$date)-min(alldata$date) + 1


```

The data was collected over a `r dur` day period. `r stepna`% of the samples were NA.

## What is mean total number of steps taken per day?

To calculate the mean number of steps per day, first calculate total steps by day for use in the histogram and use the same result to calculate the mean and median.  The plotsteps function was created to perform these calculations and make the plot

```{r Plot of steps by day}
# function returns list with mean, median, a histogram of steps per day and a count of the number of days with missing data
plotsteps<-function(data,title){

#totals by day the number of steps and the numbers of samples that were NA
  stepsperday<-ddply(data, .(date),summarize, 
                     dsteps=sum(steps),
                     nasteps = sum(is.na(steps)))
  stepsperday$date<-as.Date(stepsperday$date)  
  
#calculate mean number of steps per day
  smean<-round(mean(stepsperday$dsteps,na.rm = TRUE),0)

#calculate median number of steps per day
  smedian<-round(median(stepsperday$dsteps,na.rm = TRUE),0)

#build histogram of total steps per day 
  graph<-ggplot(stepsperday,aes(dsteps)) +
              geom_histogram(fill = "steelblue", binwidth = 500) + 
              labs(list(title = title, x="Number of Steps", y ="Number of Days"))
  
#Calculate  number of days that had NA value
  nadays<-sum(stepsperday$nasteps>0)              

out  <- list(mean = smean, median = smedian,graph = graph, NAdays = nadays)
}
```

This is a histogram of total steps by day with a bin width of 500 steps.   
```{r Histogram of Total Steps per day}
a<-plotsteps(alldata, "Total Steps per Day")
print(a$graph)

```

Ignoring the `r a$NAdays` days where no data is available, the mean number steps taken each day is `r a$mean` and the median is `r a$median`

## What is the average daily activity pattern?
First calculate the average number of steps by 5 minute period
```{r Time Series of Average daily activity pattern}

avgstepsbyperiod<-ddply(alldata,.(interval),summarize,
                        avg = mean(steps,na.rm = TRUE))

#create an vector containing interval converted into a time variable so that it can be plotted 
avgstepsbyperiod$convert<-as.difftime(str_pad(avgstepsbyperiod$interval,4,side = "left", pad= "0"),format = "%H%M")
#graph<- qplot(data = avgstepsbyperiod,x= convert,y = avg, geom = "line")
#graph< - plot(avgstepsbyperiod$convert,avgstepsbyperiod$avg, xlab = "Interval", ylab = "Average Steps",type = "l")
 graph<-ggplot(avgstepsbyperiod,aes(interval,avg)) +
                 geom_line(colour = "steelblue") + 
                 labs(list(title = "Average Daily Steps per Interval", 
                         x     = "Interval", 
                         y     = "Average Number of Steps"))
                
 print(graph)
```
Now calculate the interval with the most steps
```{r Calculate interval with most steps}
moststeps<-avgstepsbyperiod[avgstepsbyperiod$avg == max(avgstepsbyperiod$avg),]


```

The interval with the greatest average number of steps over the sample period is `r moststeps$interval` with and average of `r round(moststeps$avg,1)` steps.  

## Imputing missing values

``` {r Count Missing Values}
#Calculate the number of rows with missing data
missing<-!complete.cases(alldata)
incrows<-sum(missing)
```

To calculate the distribution of the missing data,  we count the number of missing rows by day :

```{r Calculate number of missing rows by date , results="asis"}
#Number of mssing rows by date, showing only those dates with missing rows > 0
b<-ddply(alldata,.(date),summarise, NAcount = sum(is.na(steps)))
b[b$NAcount>0,]

```
1. There are `r incrows` rows with missing data and that those rows are clustered in above listed eight days that have no data at all.
2. Values to complete these data are imputed from the mean number of steps for that 5 minute period calculated from the rest of the dataset.  Because there are eight entire days that have no data (steps = NA for every interval), it is possible to have R automatically recycle the average steps by period data calcuated above to fill the missing values into the incomplete cases
3. Now to create a new data set with the NA values replaced with the mean value for each 5-minute period
```{r Impute Missing values}
#make a copy of the original
newdata<-alldata

# put the average steps by period calculated above into the incomplete cases.  R will automatically recycle the data so that all eight days are updated
newdata[!complete.cases(newdata),"steps"]<- avgstepsbyperiod$avg

# now recalculate the number of rows with missing data
incrows<-sum(!complete.cases(newdata))
``` 

  Now `r incrows` rows are missing data

4. Now use the function defined above calculate mean and median and graph the new data set
```{r Histogram of Total Steps by  day with missing  imputed}
b<-plotsteps(newdata, "Steps per Day with Missing Values Imputed")

print(b$graph)
```             

The mean is now `r b$mean` and the median is `r b$median`.  By definition, using the 5-minute interval means to replace NAs will not change the 5-minute means and therefore the overall daily mean doesn't change because all that we have done is add eight additional days worth of data with the mean number of steps.  However we would expect the median to increase slightly by adding eight further days of data with a value greater than the median calculated with missing data. 

## Are there differences in activity patterns between weekdays and weekends?

1. Create a new factor variable identifying if day is a weekday or a weekend
```{r Create a new factor variable}

newdata[weekdays(newdata$date) %in% c("Saturday","Sunday"),"DoW"]<-"Weekend"
newdata[is.na(newdata$DoW),"DoW"]<-"Weekday"
newdata$DoW<-as.factor(newdata$DoW)


WvW<-ddply(newdata,.(DoW,interval),summarise,
                                    avg = mean(steps))
```
```{r Comparison of Activity Patterns between Weekdays and weekends}
graph<-ggplot(WvW,aes(interval,avg, group = DoW)) +
              geom_line(colour = "steelblue") + 
    #          scale_x_discrete(breaks=seq(0, 2400, by=100)) +
              labs(list(title = "Average Steps per Interval, Weekday vs Weekend", 
                        x     = "Interval", 
                        y     = "Average Number of Steps"))+
              facet_wrap(~DoW,nrow =2)
print(graph)

