---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
First read in the data from the "activity.zip" file in the working directory and convert the data in the "Date"  column into date format
```{r Load, echo = TRUE}
setwd("C:/Users/Keith_2/Documents/R/RepData_PeerAssessment1")
alldata<-read.csv(unzip("activity.zip")) 
alldata$date<-as.Date(alldata$date)


```
Here is an example of the data:
```{r example}
head(alldata,7)
```


Here is a summary of the data
```{r summary, echo = TRUE}

print(summary(alldata))
library(plyr)
library(ggplot2)
stepna<-round(100* sum(is.na(alldata$steps))/nrow(alldata),1);
dur<-max(alldata$date)-min(alldata$date) + 1
samps <- max(alldata$interval)/5 + 1
stepsperday<-ddply(alldata, .(date),summarize,
                    dsteps=sum(steps),
                    nasteps = sum(is.na(steps)))
       

nadays<-sum(stepsperday$nasteps>0)

```

The duration of the data sampling was `r dur` days with and `r nadays` of those days having no data. `r stepna`% of the samples were NA.

## What is mean total number of steps taken per day?
To calculate the mean and median number of steps taken per day, we will calculate the number the total number of steps per day and then take the mean

Ignoring days where no data is available, the mean number steps are taken each day is:

```{r Mean steps per day, echo=TRUE}

round(mean(stepsperday$dsteps,na.rm = TRUE),0)



```
And the median is :


```{r Median steps per day, echo=TRUE}

round(median(stepsperday$dsteps,na.rm = TRUE),0)



```

## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
